---
title: "Population Dynamics at HBCUs and Ivy League Institutions"
author:
  - "Ari Kamau"
  - "Nathan Alexander"
  - "Center for Applied Data Science and Analytics"
  - "Howard University"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true  # Enable floating TOC in the sidebar
    number_sections: true
    theme: cerulean
    always_allow_html: true
    self_contained: true
editor_options:
  markdown:
    wrap: sentence
---



#Setup


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
remotes::install_github("btskinner/rscorecard")
remotes::install_github('jbryer/ipeds')
install.packages("writexl")
```



```{r ipeds, include=F}
library(rscorecard)
library(ipeds)
library(ggplot2)
library(xtable)
library(writexl)
#install.packages("maps")
library(maps)
# Get your Data.gov API key at https://api.data.gov/signup/. Save your key in your R environment at the start of your R session using sc_key()
sc_key('0yRXAMFlax8c0K0rD8uJJeOwiSWrWfeXUjUfF5wn')
# list of the currently available functions
ls('package:rscorecard')
ls('package:ipeds')

# To look up information about data elements, use the sc_dict() function.
sc_dict(filter_vars = TRUE) # List of all variables
sc_dict("hi_inc_rpy_5yr_rt") # Example


```

# DATA

Loading data from a [`rscorecard`` R package](https://github.com/btskinner/rscorecard) and  [`ipeds` R package](https://github.com/jbryer/ipeds) to analyze population dynamics at HBCUs, Ivy League, and Community College Institutions.

The US Department of Education College Scorecard  contains data on higher education institutions in the United States. The data are made publically available for analysis by researchers. 

For this memo, I used the `rscorecard` and `ipeds`  package in R to perform all analysis.

##Scorecard Institutions

### HBCUs 
```{r, include=F}


HBCUs <- c(
  "220182", # Tennessee State University
  "175772", # Jackson State University
  "234155", # Virginia State University
  "176044", # Mississippi Valley State University
  "198543", # Fayetteville State University
  "198507", # Elizabeth City State University
  "199102", # North Carolina A&T State University
  "131520"  # Howard University
)

                         
SC.hbcu <- sc_init() |>  
      sc_filter(hbcu==1) |>   # Filter Data for HBCUs
    sc_select(unitid,opeid6, instnm, stabbr,    # Select Key Variables
    lo_inc_rpy_5yr_rt,
md_inc_rpy_5yr_rt,
hi_inc_rpy_5yr_rt,

d150_4,
d150_l4,
d150_4_pooled,
d150_l4_pooled,

pctpell_dcs,
pell_debt_mdn,

grad_debt_mdn_supp,
grad_debt_mdn10yr_supp,

md_earn_wne_5yr,
md_earn_wne_inc1_p6
  ) |>
    sc_year("latest") |> # Set Year
    sc_get()
SC.hbcu1<- SC.hbcu[which(SC.hbcu$unitid %in%HBCUs),]
m <- SC.hbcu1[ c("unitid",
"opeid6",
"instnm",
"stabbr",
"lo_inc_rpy_5yr_rt",
"md_inc_rpy_5yr_rt",
"hi_inc_rpy_5yr_rt",

"d150_4",
"d150_4_pooled",


"pctpell_dcs",
"pell_debt_mdn",

"grad_debt_mdn_supp",
"grad_debt_mdn10yr_supp",

"md_earn_wne_5yr",
"md_earn_wne_inc1_p6")]
names(m) <- c("Unit ID",
"OPEID","Institution",
"State",

"Low Family Income Repayment Rate (5yr)",
"Medium Family Income Repayment Rate (5yr)",
"High Family Income Repayment Rate (5yr)",

"Completion Count (150%)",

"Completion Count Rolling Average (150%)",


"Pell Percentage of DCS",
"Median Pell Debt",

"Median Debt of Completer",
"Median Debt of Completer (10yr Monthly Payment)",

"Median Earnings 5 Years After Comlpetion",
"Median Earnings 5 Years After Completion (Low Income)"

)
m



```



### Ivys 


```{r, include=F}

IvyLeague <- c(
  "186131", # Yale University
  "190150", # Columbia University
  "166027", # Cornell University
  "130794", # Dartmouth College
  "215062", # University of Pennsylvania
  "182670", # Princeton University
  "217156", # Brown University
  "190415"  # Harvard University
)

SC.ivy <- sc_init() |> 
      sc_filter(region == c(1,2), ccbasic == c(15,16)) |> # Filter Data for Ivys
    sc_select(unitid,opeid6, instnm, stabbr, # Select Key Variables
    lo_inc_rpy_5yr_rt,
    lo_inc_rpy_5yr_rt,
md_inc_rpy_5yr_rt,
hi_inc_rpy_5yr_rt,

d150_4,
d150_l4,
d150_4_pooled,
d150_l4_pooled,

pctpell_dcs,
pell_debt_mdn,

grad_debt_mdn_supp,
grad_debt_mdn10yr_supp,

md_earn_wne_5yr,
md_earn_wne_inc1_p6
  ) |>
    sc_year("latest") |> # Set Year
    sc_get()
SC.ivy1<- SC.ivy[which(SC.ivy$unitid %in%IvyLeague),]
n <- SC.ivy1[ c("unitid",
"opeid6",
"instnm",
"stabbr",
"lo_inc_rpy_5yr_rt",
"md_inc_rpy_5yr_rt",
"hi_inc_rpy_5yr_rt",

"d150_4",
"d150_4_pooled",


"pctpell_dcs",
"pell_debt_mdn",

"grad_debt_mdn_supp",
"grad_debt_mdn10yr_supp",

"md_earn_wne_5yr",
"md_earn_wne_inc1_p6")]
names(n) <- c("Unit ID",
"OPEID","Institution",
"State",

"Low Family Income Repayment Rate (5yr)",
"Medium Family Income Repayment Rate (5yr)",
"High Family Income Repayment Rate (5yr)",

"Completion Count (150%)",

"Completion Count Rolling Average (150%)",


"Pell Percentage of DCS",
"Median Pell Debt",

"Median Debt of Completer",
"Median Debt of Completer (10yr Monthly Payment)",

"Median Earnings 5 Years After Comlpetion",
"Median Earnings 5 Years After Completion (Low Income)"
  
  
  

)
n



```



### Community Colleges

```{r, include=F}

Community <- c(
"197966", # Beaufort County Community College 
"177977", # State Technical College of Missouri 
"155487", # Manhattan Area Technical College 
"114716", # Foothill College 
"163426", # Montgomery College 
"198491", # Edgecombe Community College 
"116439" # Irvine Valley College 
)

SC.comm <- sc_init() |> 
      sc_filter(region == c(2,4,5,8)) |> # Filter Data for Community Colleges
    sc_select(unitid,opeid6, instnm, stabbr, # Select Key Variables
    lo_inc_rpy_5yr_rt,
    lo_inc_rpy_5yr_rt,
md_inc_rpy_5yr_rt,
hi_inc_rpy_5yr_rt,

d150_4,
d150_l4,
d150_4_pooled,
d150_l4_pooled,

pctpell_dcs,
pell_debt_mdn,

grad_debt_mdn_supp,
grad_debt_mdn10yr_supp,

md_earn_wne_5yr,
md_earn_wne_inc1_p6
  ) |>
    sc_year("latest") |> # Set Year
    sc_get()



SC.comm1<- SC.comm[which(SC.comm$unitid %in%Community),]
o <- SC.comm1[ c("unitid",
"opeid6",
"instnm",
"stabbr",
"lo_inc_rpy_5yr_rt",
"md_inc_rpy_5yr_rt",
"hi_inc_rpy_5yr_rt",

"d150_4",
"d150_4_pooled",


"pctpell_dcs",
"pell_debt_mdn",

"grad_debt_mdn_supp",
"grad_debt_mdn10yr_supp",

"md_earn_wne_5yr",
"md_earn_wne_inc1_p6")]


names(o) <- c("Unit ID",
"OPEID","Institution",
"State",

"Low Family Income Repayment Rate (5yr)",
"Medium Family Income Repayment Rate (5yr)",
"High Family Income Repayment Rate (5yr)",

"Completion Count (150%)",

"Completion Count Rolling Average (150%)",


"Pell Percentage of DCS",
"Median Pell Debt",

"Median Debt of Completer",
"Median Debt of Completer (10yr Monthly Payment)",

"Median Earnings 5 Years After Comlpetion",
"Median Earnings 5 Years After Completion (Low Income)"
  
  
  

)
o



```



## IPEDS Institutions 


```{r, include=F}
# list of surveys that are available to download
data(surveys)
surveys[,c('SurveyID', 'Title')]

# list of all available CIP codes
data(cipcodes)
names(cipcodes)

# data available
ipeds::available_ipeds() 
```




```{r, eval=F}
# sets machine timeout time 
options(timeout = 600)

#("ipeds.download.dir" = "/Users/Ari/Desktop/Quant Shop/College Default Rate/ipeds/data/downloaded/")

#ipeds::download_ipeds(
 #year = 2021, 
 #dir = "~/Desktop/Quant Shop/College Default Rate/ipeds/data/downloaded",
 #force=T)  


ipeds2021 <- ipeds::load_ipeds(2021) # load data
names(ipeds2021)



hd2021 <- ipeds::ipeds_survey(table = 'HD', year = 2021)
names(hd2021)
names(hd2021) <- tolower(names(hd2021))
hd2021 <- ipeds::recodeDirectory(hd2021)

HBCUs <- c(
  "220182", # Tennessee State University
  "175772", # Jackson State University
  "234155", # Virginia State University
  "176044", # Mississippi Valley State University
  "198543", # Fayetteville State University
  "198507", # Elizabeth City State University
  "199102", # North Carolina A&T State University
  "131520"  # Howard University
)

hd2021.hbcu <- hd2021[which(hd2021$unitid %in%HBCUs),]
p <- hd2021.hbcu[, c("instnm", "webaddr", "stabbr", "control","unitid")]
names(p) <- c("Institution", "Web Address", "State", "Sector","Unit ID")
p

IvyLeague <- c(
  "186131", # Yale University
  "190150", # Columbia University
  "166027", # Cornell University
  "130794", # Dartmouth College
  "215062", # University of Pennsylvania
  "182670", # Princeton University
  "217156", # Brown University
  "190415"  # Harvard University
)

hd2021.ivy <- hd2021[which(hd2021$unitid %in%IvyLeague),]
q <- hd2021.ivy[, c("instnm", "webaddr", "stabbr", "control","unitid")]
names(q) <- c("Institution", "Web Address", "State", "Sector","Unit ID" )
q

Community <- c(
"197966", # Beaufort County Community College 
"177977", # State Technical College of Missouri 
"155487", # Manhattan Area Technical College 
"114716", # Foothill College 
"163426", # Montgomery College 
"198491", # Edgecombe Community College 
"116439" # Irvine Valley College 
)

hd2021.comm <- hd2021[which(hd2021$unitid %in%Community),]
r <- hd2021.comm[, c("instnm", "webaddr", "stabbr", "control","unitid")]
names(r) <- c("Institution", "Web Address", "State", "Sector","Unit ID")
r

```



# Merge Scorecard and IPEDS Data
 
Note: For future reference, we may only use Unit Id ( 6 digit OPE ID ) for merging as it is the most compact and translatable variable across data sets.
```{r, include=F}

# Merge each Institutions Data using Key Variable
MP.HBCU<- merge(m, p, by = c("Unit ID","Institution","State"), all = TRUE)
NQ.IVY<- merge(n, q, by = c("Unit ID","Institution","State"), all = TRUE)
OR.COMM<- merge(o, r, by = c("Unit ID","Institution","State"), all = TRUE)


# Combine into a named list
Inital_Default_List  <- list(
"Initial HBCU Default Data" = MP.HBCU, 
"Initial Ivy Default Data" = NQ.IVY,
"Initial Community Default Data" = OR.COMM
)
# Export to one file with multiple sheets
write_xlsx(Inital_Default_List, "Inital_Default.xlsx")
```

