---
title: "Population Dynamics at HBCUs and Ivy League Institutions"
author:
  - "Ari Kamau"
  - "Nathan Alexander"
  - "Center for Applied Data Science and Analytics"
  - "Howard University"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true  # Enable floating TOC in the sidebar
    number_sections: true
    theme: cerulean
    always_allow_html: true
    self_contained: true
editor_options:
  markdown:
    wrap: sentence
---



#Setup


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
remotes::install_github("btskinner/rscorecard")
remotes::install_github('jbryer/ipeds')
install.packages("writexl")
install.packages("here")
library(here)
here::i_am("hbcu-ivy.Rmd")   # Uses the "here" R package to set the current RMD file as the top directory (Easier Reproduceability)
```



```{r ipeds, include=F}
library(rscorecard)
library(ipeds)
library(ggplot2)
library(xtable)
library(writexl)
#install.packages("maps")
library(maps)

# Get your Data.gov API key at https://api.data.gov/signup/. Save the key in your R environment at the start of your R session using sc_key()
sc_key('0yRXAMFlax8c0K0rD8uJJeOwiSWrWfeXUjUfF5wn')
# list of the currently available functions
ls('package:rscorecard')
ls('package:ipeds')

# To look up information about data elements, use the sc_dict() function.
sc_dict(filter_vars = TRUE) # List of all variables
sc_dict("c150_4") # Example

```

# DATA

Loading data from a [`rscorecard`` R package](https://github.com/btskinner/rscorecard) and  [`ipeds` R package](https://github.com/jbryer/ipeds) to analyze population dynamics at HBCUs, Ivy League, and Community College Institutions.

The US Department of Education College Scorecard  contains data on higher education institutions in the United States. The data are made publically available for analysis by researchers. 

For this memo, I used the `rscorecard` and `ipeds`  package in R to perform all analysis.

##Scorecard Institutions

### HBCUs 
```{r, include=F}


HBCUs <- c(
  "220182", # Tennessee State University
  "175772", # Jackson State University
  "234155", # Virginia State University
  "176044", # Mississippi Valley State University
  "198543", # Fayetteville State University
  "198507", # Elizabeth City State University
  "199102", # North Carolina A&T State University
  "131520"  # Howard University
)

                         
SC.hbcu <- sc_init() |>  
      sc_filter(hbcu==1) |>   # Filter Scorecard Data for HBCUs
    sc_select(                # Select Key Variables
unitid,                 # Unique 6-digit identifier for each institution (Department of Education's National Center for Education Statistics (NCES))
opeid6,                 # 6-digit OPE ID; used to link to federal aid/NSLDS data
instnm,                 # Institution name
stabbr,                 # State postal abbreviation
lo_inc_rpy_5yr_rt,      # 5-year repayment rate for low-income borrowers (family income <$30,000)
md_inc_rpy_5yr_rt,      # 5-year repayment rate for middle-income borrowers (family income $30,000–$75,000)
hi_inc_rpy_5yr_rt,      # 5-year repayment rate for high-income borrowers (family income >$75,000)
c150_4,                 # Adjusted cohort rate for 150% completion rate at 4-year institutions
d150_4,                 # Adjusted cohort count (denominator) for 150% completion rate at 4-year institutions
pctpell_dcs,            # Share of degree/certificate-seeking undergraduates awarded a Pell Grant
pell_debt_mdn,          # Median cumulative loan debt for Pell Grant recipients at time of leaving institution
grad_debt_mdn_supp,     # Median cumulative loan debt for completers, suppressed when n<30
grad_debt_mdn10yr_supp, # Median completer debt expressed as estimated monthly payment under 10-year amortization, suppressed when n<30
md_earn_wne_5yr,        # Median earnings of graduates working and not enrolled 5 years after completion 
md_earn_wne_inc1_p6     # Median earnings 6 years after entry for students in the lowest earnings tercile
) |>
    sc_year("latest") |> # Set Year
    sc_get()
SC.hbcu1<- SC.hbcu[which(SC.hbcu$unitid %in%HBCUs),] # Filter for Key Variables at selcted HBCUs
m <- SC.hbcu1[ c(        # Rename HBCU Data set "m"
vars <- c(               # Restate key variables          
"unitid",                 # Unique 6-digit identifier for each institution (Department of Education's National Center for Education Statistics (NCES))
"opeid6",                 # 6-digit OPE ID; used to link to federal aid/NSLDS data
"instnm",                 # Institution name
"stabbr",                 # State postal abbreviation
"lo_inc_rpy_5yr_rt",      # 5-year repayment rate for low-income borrowers (family income <$30,000)
"md_inc_rpy_5yr_rt",      # 5-year repayment rate for middle-income borrowers (family income $30,000–$75,000)
"hi_inc_rpy_5yr_rt",      # 5-year repayment rate for high-income borrowers (family income >$75,000)
"c150_4",                 # Adjusted cohort rate for 150% completion rate at 4-year institutions institutions; from IPEDS Graduation Rates component
"d150_4",                 # Adjusted cohort count (denominator) for 150% completion rate at 4-year  institutions
"pctpell_dcs",            # Share of degree/certificate-seeking undergraduates awarded a Pell Grant; 
"pell_debt_mdn",          # Median cumulative loan debt for Pell Grant recipients at time of leaving institution
"grad_debt_mdn_supp",     # Median cumulative loan debt for completers, suppressed when n<30
"grad_debt_mdn10yr_supp", # Median completer debt expressed as estimated monthly payment under 10-year amortization, suppressed when n<30
"md_earn_wne_5yr",        # Median earnings of graduates working and not enrolled 5 years after completion
"md_earn_wne_inc1_p6"     # Median earnings 6 years after entry for students in the lowest earnings tercile
))]

names(m) <- c(   # Renamed key variables     
"Unit ID",                                         # unitid
"OPEID",                                           # opeid6
"Institution",                                     # instnm
"State",                                           # stabbr
"Low Family Income Repayment Rate (5yr)",          # lo_inc_rpy_5yr_rt
"Medium Family Income Repayment Rate (5yr)",       # md_inc_rpy_5yr_rt
"High Family Income Repayment Rate (5yr)",         # hi_inc_rpy_5yr_rt
"Completion Rate (150%)",                          # c150_4
"Completion Count (150%)",                         # d150_4
"Pell Percentage of DCS",                          # pctpell_dcs
"Median Pell Debt",                                # pell_debt_mdn
"Median Debt of Completer",                        # grad_debt_mdn_supp
"Median Debt of Completer (10yr Monthly Payment)", # grad_debt_mdn10yr_supp
"Median Earnings 5 Years After Comlpetion",        # md_earn_wne_5yr
"Median Earnings 6 Years After Entry (Low Income)" # md_earn_wne_inc1_p6
)
m



```



### Ivys 


```{r, include=F}

IvyLeague <- c(
  "186131", # Yale University
  "190150", # Columbia University
  "166027", # Cornell University
  "130794", # Dartmouth College
  "215062", # University of Pennsylvania
  "182670", # Princeton University
  "217156", # Brown University
  "190415"  # Harvard University
)

SC.ivy <- sc_init() |> 
   sc_filter(region == c(1,2), ccbasic == c(15,16)) |> # Filter Scorecard Data for Ivys
   sc_select(                                          # Select Key Variables
unitid,                 # Unique 6-digit identifier for each institution (Department of Education's National Center for Education Statistics (NCES))
opeid6,                 # 6-digit OPE ID; used to link to federal aid/NSLDS data
instnm,                 # Institution name
stabbr,                 # State postal abbreviation
lo_inc_rpy_5yr_rt,      # 5-year repayment rate for low-income borrowers (family income <$30,000)
md_inc_rpy_5yr_rt,      # 5-year repayment rate for middle-income borrowers (family income $30,000–$75,000)
hi_inc_rpy_5yr_rt,      # 5-year repayment rate for high-income borrowers (family income >$75,000)
c150_4,                 # Adjusted cohort rate for 150% completion rate at 4-year institutions
d150_4,                 # Adjusted cohort count (denominator) for 150% completion rate at 4-year institutions
pctpell_dcs,            # Share of degree/certificate-seeking undergraduates awarded a Pell Grant
pell_debt_mdn,          # Median cumulative loan debt for Pell Grant recipients at time of leaving institution
grad_debt_mdn_supp,     # Median cumulative loan debt for completers, suppressed when n<30
grad_debt_mdn10yr_supp, # Median completer debt expressed as estimated monthly payment under 10-year amortization, suppressed when n<30
md_earn_wne_5yr,        # Median earnings of graduates working and not enrolled 5 years after completion 
md_earn_wne_inc1_p6     # Median earnings 6 years after entry for students in the lowest earnings tercile
  ) |>
    sc_year("latest") |> # Set Year
    sc_get()
SC.ivy1<- SC.ivy[which(SC.ivy$unitid %in%IvyLeague),] # Filter for Key Variables at selcted HBCUs
n <- SC.ivy1[ c(        # Rename HBCU Data set "m"
vars <- c(               # Restate key variables          
"unitid",                 # Unique 6-digit identifier for each institution (Department of Education's National Center for Education Statistics (NCES))
"opeid6",                 # 6-digit OPE ID; used to link to federal aid/NSLDS data
"instnm",                 # Institution name
"stabbr",                 # State postal abbreviation
"lo_inc_rpy_5yr_rt",      # 5-year repayment rate for low-income borrowers (family income <$30,000)
"md_inc_rpy_5yr_rt",      # 5-year repayment rate for middle-income borrowers (family income $30,000–$75,000)
"hi_inc_rpy_5yr_rt",      # 5-year repayment rate for high-income borrowers (family income >$75,000)
"c150_4",                 # Adjusted cohort rate for 150% completion rate at 4-year institutions institutions; from IPEDS Graduation Rates component
"d150_4",                 # Adjusted cohort count (denominator) for 150% completion rate at 4-year  institutions
"pctpell_dcs",            # Share of degree/certificate-seeking undergraduates awarded a Pell Grant; 
"pell_debt_mdn",          # Median cumulative loan debt for Pell Grant recipients at time of leaving institution
"grad_debt_mdn_supp",     # Median cumulative loan debt for completers, suppressed when n<30
"grad_debt_mdn10yr_supp", # Median completer debt expressed as estimated monthly payment under 10-year amortization, suppressed when n<30
"md_earn_wne_5yr",        # Median earnings of graduates working and not enrolled 5 years after completion
"md_earn_wne_inc1_p6"     # Median earnings 6 years after entry for students in the lowest earnings tercile
))]

names(n) <- c(   # Renamed key variables     
"Unit ID",                                         # unitid
"OPEID",                                           # opeid6
"Institution",                                     # instnm
"State",                                           # stabbr
"Low Family Income Repayment Rate (5yr)",          # lo_inc_rpy_5yr_rt
"Medium Family Income Repayment Rate (5yr)",       # md_inc_rpy_5yr_rt
"High Family Income Repayment Rate (5yr)",         # hi_inc_rpy_5yr_rt
"Completion Rate (150%)",                          # c150_4
"Completion Count (150%)",                         # d150_4
"Pell Percentage of DCS",                          # pctpell_dcs
"Median Pell Debt",                                # pell_debt_mdn
"Median Debt of Completer",                        # grad_debt_mdn_supp
"Median Debt of Completer (10yr Monthly Payment)", # grad_debt_mdn10yr_supp
"Median Earnings 5 Years After Comlpetion",        # md_earn_wne_5yr
"Median Earnings 6 Years After Entry (Low Income)" # md_earn_wne_inc1_p6
)
n


```



### Community Colleges

```{r, include=F}

Community <- c(
"197966", # Beaufort County Community College 
"177977", # State Technical College of Missouri 
"155487", # Manhattan Area Technical College 
"114716", # Foothill College 
"163426", # Montgomery College 
"198491", # Edgecombe Community College 
"116439" # Irvine Valley College 
)

SC.comm <- sc_init() |> 
      sc_filter(region == c(2,4,5,8)) |> # Filter Data for Community Colleges
    sc_select(                                          # Select Key Variables
unitid,                 # Unique 6-digit identifier for each institution (Department of Education's National Center for Education Statistics (NCES))
opeid6,                 # 6-digit OPE ID; used to link to federal aid/NSLDS data
instnm,                 # Institution name
stabbr,                 # State postal abbreviation
lo_inc_rpy_5yr_rt,      # 5-year repayment rate for low-income borrowers (family income <$30,000)
md_inc_rpy_5yr_rt,      # 5-year repayment rate for middle-income borrowers (family income $30,000–$75,000)
hi_inc_rpy_5yr_rt,      # 5-year repayment rate for high-income borrowers (family income >$75,000)
c150_4,                 # Adjusted cohort rate for 150% completion rate at 4-year institutions
d150_4,                 # Adjusted cohort count (denominator) for 150% completion rate at 4-year institutions
pctpell_dcs,            # Share of degree/certificate-seeking undergraduates awarded a Pell Grant
pell_debt_mdn,          # Median cumulative loan debt for Pell Grant recipients at time of leaving institution
grad_debt_mdn_supp,     # Median cumulative loan debt for completers, suppressed when n<30
grad_debt_mdn10yr_supp, # Median completer debt expressed as estimated monthly payment under 10-year amortization, suppressed when n<30
md_earn_wne_5yr,        # Median earnings of graduates working and not enrolled 5 years after completion 
md_earn_wne_inc1_p6     # Median earnings 6 years after entry for students in the lowest earnings tercile 
  ) |>
    sc_year("latest") |> # Set Year
    sc_get()
SC.comm1<- SC.comm[which(SC.comm$unitid %in%Community),] # Filter for Key Variables at selcted HBCUs
o <- SC.comm1[ c(        # Rename HBCU Data set "m"
vars <- c(               # Restate key variables          
"unitid",                 # Unique 6-digit identifier for each institution (Department of Education's National Center for Education Statistics (NCES))
"opeid6",                 # 6-digit OPE ID; used to link to federal aid/NSLDS data
"instnm",                 # Institution name
"stabbr",                 # State postal abbreviation
"lo_inc_rpy_5yr_rt",      # 5-year repayment rate for low-income borrowers (family income <$30,000)
"md_inc_rpy_5yr_rt",      # 5-year repayment rate for middle-income borrowers (family income $30,000–$75,000)
"hi_inc_rpy_5yr_rt",      # 5-year repayment rate for high-income borrowers (family income >$75,000)
"c150_4",                 # Adjusted cohort rate for 150% completion rate at 4-year institutions institutions; from IPEDS Graduation Rates component
"d150_4",                 # Adjusted cohort count (denominator) for 150% completion rate at 4-year  institutions
"pctpell_dcs",            # Share of degree/certificate-seeking undergraduates awarded a Pell Grant; 
"pell_debt_mdn",          # Median cumulative loan debt for Pell Grant recipients at time of leaving institution
"grad_debt_mdn_supp",     # Median cumulative loan debt for completers, suppressed when n<30
"grad_debt_mdn10yr_supp", # Median completer debt expressed as estimated monthly payment under 10-year amortization, suppressed when n<30
"md_earn_wne_5yr",        # Median earnings of graduates working and not enrolled 5 years after completion
"md_earn_wne_inc1_p6"     # Median earnings 6 years after entry for students in the lowest earnings tercile
))]

names(o) <- c(   # Renamed key variables     
"Unit ID",                                         # unitid
"OPEID",                                           # opeid6
"Institution",                                     # instnm
"State",                                           # stabbr
"Low Family Income Repayment Rate (5yr)",          # lo_inc_rpy_5yr_rt
"Medium Family Income Repayment Rate (5yr)",       # md_inc_rpy_5yr_rt
"High Family Income Repayment Rate (5yr)",         # hi_inc_rpy_5yr_rt
"Completion Rate (150%)",                          # c150_4
"Completion Count (150%)",                         # d150_4
"Pell Percentage of DCS",                          # pctpell_dcs
"Median Pell Debt",                                # pell_debt_mdn
"Median Debt of Completer",                        # grad_debt_mdn_supp
"Median Debt of Completer (10yr Monthly Payment)", # grad_debt_mdn10yr_supp
"Median Earnings 5 Years After Comlpetion",        # md_earn_wne_5yr
"Median Earnings 6 Years After Entry (Low Income)" # md_earn_wne_inc1_p6
)
o


```



## IPEDS Institutions 


```{r, include=F}
# list of surveys that are available to download
data(surveys)
surveys[,c('SurveyID', 'Title')]

# list of all available CIP codes
data(cipcodes)
names(cipcodes)

# data available
ipeds::available_ipeds() 
```




```{r, eval=F}
# sets machine timeout time 
options(timeout = 600)


#Download the IPEDS data
ipeds::download_ipeds(
 year = 2021,             # Set Year
 dir = "data/downloaded", # using the here::i_am function modifies machine specific working directory pathways into project specific pathways improving reproducibility
 force=TRUE)              

ipeds2021 <- ipeds::load_ipeds(2021) # load data
names(ipeds2021)



hd2021 <- ipeds::ipeds_survey(table = 'HD', year = 2021)
names(hd2021)
names(hd2021) <- tolower(names(hd2021))
hd2021 <- ipeds::recodeDirectory(hd2021)

HBCUs <- c(
  "220182", # Tennessee State University
  "175772", # Jackson State University
  "234155", # Virginia State University
  "176044", # Mississippi Valley State University
  "198543", # Fayetteville State University
  "198507", # Elizabeth City State University
  "199102", # North Carolina A&T State University
  "131520"  # Howard University
)

hd2021.hbcu <- hd2021[which(hd2021$unitid %in%HBCUs),]
p <- hd2021.hbcu[, c( # Key Variable Selection
"instnm",  # Institution
"webaddr", # Web address of the School
"stabbr",  # State 2 letter Abbreviation 
"control", # Institution Sector
"unitid"   # 6-digit identification number (NCES)
)]
names(p) <- c("Institution", "Web Address", "State", "Sector","Unit ID")


IvyLeague <- c(
  "186131", # Yale University
  "190150", # Columbia University
  "166027", # Cornell University
  "130794", # Dartmouth College
  "215062", # University of Pennsylvania
  "182670", # Princeton University
  "217156", # Brown University
  "190415"  # Harvard University
)

hd2021.ivy <- hd2021[which(hd2021$unitid %in%IvyLeague),]
q <- hd2021.ivy[, c(
"instnm",  # Institution
"webaddr", # Web address of the School
"stabbr",  # State 2 letter Abbreviation 
"control", # Institution Sector
"unitid"   # 6-digit identification number (NCES)
)]
names(q) <- c("Institution", "Web Address", "State", "Sector","Unit ID" )
q

Community <- c(
"197966", # Beaufort County Community College 
"177977", # State Technical College of Missouri 
"155487", # Manhattan Area Technical College 
"114716", # Foothill College 
"163426", # Montgomery College 
"198491", # Edgecombe Community College 
"116439" # Irvine Valley College 
)

hd2021.comm <- hd2021[which(hd2021$unitid %in%Community),]
r <- hd2021.comm[, c(
"instnm",  # Institution
"webaddr", # Web address of the School
"stabbr",  # State 2 letter Abbreviation 
"control", # Institution Sector
"unitid"   # 6-digit identification number (NCES)
)]

names(r) <- c("Institution", "Web Address", "State", "Sector","Unit ID")
r

```



# Merge Scorecard and IPEDS Data
 
Note: For future reference, we may only use `Unit ID` ( 6 digit OPE ID ) for merging as it is the most compact and translatable variable across data sets. `Institution` and `State` are included to remove repeated columns. 


```{r, include=F}

# Merge each Institutions Data using Unit ID as Key Variable
MP.HBCU<- merge(m, p, by = c("Unit ID","Institution","State"), all = TRUE) 
NQ.IVY<- merge(n, q, by = c("Unit ID","Institution","State"), all = TRUE)
OR.COMM<- merge(o, r, by = c("Unit ID","Institution","State"), all = TRUE)


# Combine into a named list
Inital_Default_List  <- list(
"Initial HBCU Default Data" = MP.HBCU, 
"Initial Ivy Default Data" = NQ.IVY,
"Initial Community Default Data" = OR.COMM
)
# Export to one file with multiple sheets
write_xlsx(Inital_Default_List, "Inital_Default.xlsx")
```

